---
title: "QMSS R Tutorial Part 5: Missing values"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# NA

A common problem when analysing data is dealing with missing data.  

The variable below is a vector with two missing values, which is a *logical constant* which represents "Not available".

```{r}
x = c(2, 5, 3, NA, 1, NA)
```

> **Task**: What happens if you try to calculate the sum of this variable?

You get NA returned!  This isn't helpful, but is technically correct - there's no defined sum for missing data.

There are two ways of dealing with this.  First, take out NA values, and we can do this by indexing **x**.  However, you can't use the boolean operators with NAs:

```{r}
x == NA
```

Instead, you can use the function `is.na`:

```{r}
is.na(x)
```

We want to keep only the values that are **not** NA, so we can negate the booleans with the symbol `!`:

```{r}
!is.na(x)
x2 = x[!is.na(x)]
sum(x2)
```

The second option is to use the optional argument `na.rm` in `sum`.  This argument controls whether NA values should be ignored when calcualting the sum.

```{r}
sum(x, na.rm=TRUE)
```

When using this, be sure you know what you are excluding.

NA values can also be problematic when indexing:

```{r}
# a vector of numbers with 1 NA
x = c(1,2,NA,4)
# a vector of characters
y = c("a",'b','c','d')
# index y using x
y2 = y[x]
y2
```

Suddenly, y2 includes NA values!

# Data with NA values

In the previous data sets we looked at, there were no missing data.  The file *WALS_ObjectVerbOrder.csv* does contain missing values.  Let's load it:

```{r echo=c(2:5)}
setwd("~/Documents/Teaching/JenaSpringSchool/org/spring-school/IntroToR")

d = read.csv("data/WALS_ObjectVerbOrder.csv", stringsAsFactors = F)
dim(d)
head(d)
```

We see that some of the data for the variable **ObjectVerbOrder** is missing, and also some glotto and iso codes.

> **Task**: How many datapoints in ObjectVerbOrder are missing?  How many glottocodes are missing?  Remember that `is.na` creates a vector of booleans and `sum` can count the number of `TRUE` values in a vector.

```{r include=F}
sum(is.na(d$ObjectVerbOrder))
sum(is.na(d$glottocode))
```

Missing values in fields which are used for indexing, like *glottocode* can be problematic.  For example, suppose we want to get the **ObjectVerbOrder** of the lagnuage Aari with glottocode "aari1239":

```{r}
d[d$glottocode=="aari1239", ]$ObjectVerbOrder
```

Even though there's only 1 data point with the glottocode "aari1239", we still get a list of 8 objects.  In fact, we get an NA returned for every NA glottocode.

There are two ways to handle this.  Either remove the NAs *in situ* using `is.na`, or remove NAs from the data frame.  The former takes up less memory, but look messier.

```{r}
# in situ:
d[!is.na(d$glottocode) & d$glottocode=="aari1239", ]$ObjectVerbOrder

# remove NAs:
d2 <- d[!is.na(d$glottocode),]
d2[d2$glottocode=="aari1239",]$ObjectVerbOrder
```


# NaN and Inf

Sometimes, calculatings can lead to unexpected results.

For example, if we divide the longitude by the latitude, what is the maximum number?

```{r}
longByLat <- d$longitude / d$latitude
head(longByLat)
max(longByLat)
```

We get `Inf` - which represents infinity.  That's because some latitude values are 0, and when we divide any number by 0, R returns infinity.

Other possible results include 'NaN' which stands for "not a number".  This can happen when a calculation has an undefined result:

```{r}
0/0
```


You now know enough to get this joke:

```{r results="hide"}
x = c(1,2)
c(x[3:18], "Batman")
```


---

Go to the [next tutorial](IntroToR_6.html)

Back to the [index](../IntroToR.html)