---
title: "QMSS R Tutorial: Answers to tasks"
output: html_document
---


# Part 1

> **Task**: can you apply the functions `sum` and `length` to boolean vectors?

Yes - sum counts the number of `TRUE` values.


> **Task**: make a list of booleans where the items are `TRUE` if each corresponding item of `nums` is less or equal to 3

```{r eval=F}
nums <= 3
```

> **Task**: make a list of booleans where the items are `TRUE` if each corresponding item of `nums` is either 5 or 1.

```{r eval=F}
nums %in% c(5,1)
```

> **Task **: Using y to index `nums` worked because the length of the two vectors are the same.  What happens if you make another variable `y2` which only has 3 boolean items, and try to index `nums` using that?

```{r eval=F}
y2 = c(TRUE, FALSE, TRUE)
nums[y2]
```


> **Task**:  Find the favourite numbers of all my friends who have more than 4 characters in their name.
> 
> Hint: build this up step by step.  First of all, get the number of characters in each name, then test whether this number is greater than 4.  This should result in a vector of booleans.  Then index nums using this vector.

```{r eval=F}
nums[nchar(friends)>4]
```


# Part 3

> **Task**: What happens if you type `d` into the console to see what's inside the object `d`?

You get a lot of data printed to the screen!  Very unhelpful.


> **Task**: Make a boolean vector which is `TRUE` if the latitude of a language is greater than 0, and `FALSE` otherwise.  Assign this to a variable named `northernHemisphere'.

```{r eval=F}
northernHemisphere <- d$latitude > 0
```


> **Task**: Make a table of counts of basic word order types for languages in the northern hemisphere.  *Hint*: you should index the rows with the variable northernHemisphere and the **BasicWordOrder** column.

```{r eval=F}
table(d[northernHemisphere,]$BasicWordOrder)
```


> **Task**: Load the data in `data/Glottolog_Data.csv` into a data frame called glottoData.

```{r eval=FALSE}
glottoData <- read.csv("data/Glottolog_Data.csv", stringsAsFactors = F)
```

> **Task** What are the names and formats of the variables in glottoData?

```{r eval=FALSE}
names(glottoData)
```

> **Task**:  Could you match the databases on the name of the language?  Use `d2` to find how many languages have different names in Glottolog versus WALS.

```{r eval=F}
sum(d2$Name!=d2$glotto.name)
```


# Part 4

> **Task**:  Which basic word order type has the higest mean latitude?


```{r eval=F}
langFamLatitudeMEAN<- tapply(d2$latitude, d2$glotto.family, mean)
langFamLatitudeMEAN.sorted <- sort(langFamLatitudeMEAN, decreasing = TRUE)
head(langFamLatitudeMEAN.sorted)
```


# Part 5

> **Task**: What happens if you try to calculate the sum of this variable?

You get NA returned!


> **Task**: How many datapoints in ObjectVerbOrder are missing?  How many glottocodes are missing?  Remember that `is.na` creates a vector of booleans and `sum` can count the number of `TRUE` values in a vector.

```{r eval=F}
sum(is.na(d$ObjectVerbOrder))
sum(is.na(d$glottocode))
```