---
title: "QMSS R Tutorial Part 6: if, for and lists"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

R includes some basic programming tools like `if` and `for` statements, and another data structure called a "list".  These aren't always necessary, but sometimes come in helpful.

# If

If statements run a piece of code only if a statement evaluates to `TRUE`.

In the code below, we pick 5 random numbers (from the normal distribution, mean of 0 and standard deviation of 1), then "print" something to the console if the first number is bigger than zero.


```{r}
# First, we set a random seed, so that my results are the same as yours.
set.seed(992)
# choose 5 random numbers
x <- rnorm(5)
# take a look
x
# is the first bigger than 0?
x[1] > 0

if(x[1] > 0){
  x[1] <- x * 1000
}

# take a look at x again
x

```

We can also optionally include a bit of code to run if the statement is `FALSE`.  Here we test the 5th number:

```{r}
if(x[5] > 0){
  print("First number is bigger than 0")
} else{
  print("First number is not bigger than 0")
}
```

# For

A `for` loop is a statement which creates a vector of variables, then runs a bit of code .

```{r}
# Make a list of numbers from 1 to 10
myVector <- 1:10
for(i in myVector){
  # double the number
  doublei <- i * 2
  # print the number
  print(doublei)
}
```

We read this as "for each i in myVector, do the following: double i , then print that value."

The variable `i` is not defined before the for loop. Instead, `i` is set to the first item in the vector, then the code within the for loop is run with `i` having that value.  Then `i` is given the next value in the vector, and the code is run again.

While `i` is still accessible outside the for loop, it is best practice to pretend that `i` only exists inside the loop.

Note that you can name the variable anything, not just `i`.

This is sometimes useful when we want to make a new list.  For example, if we want to add numbers to a vector from another vector, but only if they are greater than zero.

```{r}

set.seed(992)
# choose 5 random numbers
x <- rnorm(5)

# create empty vector for positive numbers
posNums <- c()

# for each item in x
for(i in x){
  # if the item is more than zero,
  if(i > 0){
    # add the item to the list posNums
    # (add a number to the list by using 'c')
    posNums <- c(posNums,i)
  }
}

# Look at x
x

# Look at posNums
posNums

```

However, in R there is almost always an easier and faster way to do things than using a for loop.  For example, the code below does the same thing:

```{r}
set.seed(992)
# choose 5 random numbers
x <- rnorm(5)
# use indexing to get positive numbers
posNums = x[x>0]

x
posNums
```

# Lists

Previously, we've been using vectors of numbers (1 dimensional) and data frames (2 dimensional).  This is usually enough for most needs, but R also has a data structure called a `list`.  Lists work in a similar way to vectors, but the items inside them can be vectors themselves, and of varying lengths.

Imagine we have 3 students who have sat various exams.  In the code below, we define an empty list, then add three vectors of scores to it (one for each student):

```{r}
scores = list()
scores[[1]] <- c(76, 86, 90)
scores[[2]] <- c(70, 88)
scores[[3]] <- c(50, 59, 72, 66)
```

We can now access members of the list with double square bracksts:

```{r}
scores[[2]]
```

Or use them in a function like `lapply`, which applies a function to each member of the list, then returns the result.

```{r}
lapply(scores, mean)
```

This returns a list where each member of the list contains the mean score for the student.




---

Go to the [next tutorial](IntroToR_7.html)

Back to the [index](../IntroToR.html)